---
title: Parameter estimation using randomized starting values for the polio case study
author: Aaron A. King and Edward L. Ionides
output:
  html_document:
    toc: no
    includes:
      after_body:
      - ../_includes/supp_bottom.html
      - ../_includes/license.html
bibliography: ../sbied.bib
csl: ../jss.csl
---
	
-----------------------------------

```{r knitr-opts,include=FALSE,purl=FALSE}
source("../_includes/setup.R", local = knitr::knit_global())
```
```{r prelims,include=FALSE,cache=FALSE}
set.seed(594709947L)
library(pomp)
```
 
### Exercise

Think about possible improvements on the assignment of randomized starting values for the parameter estimation searches. Propose and try out a modification of the procedure. Does it make a difference?

---------
	
### Solution

The starting points are determined by the specification of the region in which they are chosen, and how they are selected within this region.
We have followed a fairly simple and natural choice to make independent uniform draws from a specified box.
`profileDesign()` also supports a pseudo-random Sobol design rather than independent draws.
Whether a specified box has a reasonable extent can be checked by plotting the starting values on a scatterplot with the candidate maximizations obtained, as demonstrated in the [global search example](../mif) from the [iterated filtering tutorial](../mif).

The transformed scale may be more reasonable for selecting uniformly distributed starting values. For example, if we are not sure about the order of magnitude of a non-negative parameter, and we set its box interval to $[0.1,10]$, we likely want to spend half our search effort with initial values in the interval $[0.1,1]$. Sampling unformly on the log scale achieves this. 
We can change our code as follows to implement this for the polio profile calculation.

	
```{r eval=FALSE}
idx <- colnames(box)!="rho"
profile_design(
  rho=seq(0.01,0.025,length=polio_profile_pts),
  lower=box["min",idx],upper=box["max",idx],
  nprof=polio_profile_Nreps
) -> starts
```
becomes
		
```{r eval=FALSE}
idx <- which(colnames(box)!="rho")
profile_design(
  rho=seq(0.01,0.025,length=polio_profile_pts),
  lower=partrans(polio,box["min",],dir="toEst")[idx],     
  upper=partrans(polio,box["max",],dir="toEst")[idx],
  nprof=polio_profile_Nreps
) |>
starts <- data.frame(t(partrans(polio,t(trans_starts),dir="fromEst")))
```
 
```{r include=FALSE}
library(tidyverse)
read_csv("params.csv") |>
  filter(logLik_se < 1) |>
  filter(logLik==max(logLik)) |>
  distinct() -> mle1
read_csv("starting-values-exercise/params.csv") |>
  filter(logLik_se < 1) |>
  filter(logLik==max(logLik)) |>
  distinct() -> mle2
```
 
A comparable change was also made for the global search. The results from running the modified version are at [starting-values-exercise/main.Rout.save](starting-values-exercise/main.Rout.save).
The maximized log likelihood of $`r round(mle2$logLik,1)`$ is no substantial improvement on the previous value of $`r round(mle1$logLik,1)`$.
`r if (mle2$logLik-mle1$logLik>1) "**The last statement is a lie!**."`
However, the pairs plot of estimates from the global search does reveal an advantage for the modified scale.
The distinct mode with low $\sigma_{\mathrm{env}}$ is better explored with the modified starting value distribution.
When making uniform draws on an untransformed scale, very few initial values explore a region where a parameter has small order of magnitude.
In this case, the small $\sigma_{\mathrm{env}}$ mode has likelihood around 15 log units short of the maximum so it is not a competitive explanation of the data.
However, in some other situation, this shows how transforming the starting value distribution could be useful.

```{r profile_rho_plot1,echo=FALSE,fig.width=6,fig.height=6,out.width="70%"}
read_csv("starting-values-exercise/params.csv") |>
  filter(logLik>max(logLik)-10) -> ests
pairs(~logLik+psi+rho+tau+sigma_dem+sigma_env,data=subset(ests,logLik>max(logLik)-10))
```

----------------------------------

This version produced in R `r getRversion()` on `r format(Sys.Date(), "%B %d, %Y")`.
